Columnar VCF for Biobank scale datasets using Zarr

### 2024-04-09 paper.tex
Some narrative

\subsection{Computing on the genotype matrix}
Compressing data isn't ultimately the point - we want to use it.
We discuss the problem of running computations on the entire
genotype matrix, and how different formats limit what can be done.

[Review computational accessibility of the available formats.
We have those that only produce VCF in response to limited 
query languages. Those with APIs.]

We illustrate the discussion by the afdist operation from
the bcftools plugins package. This is perhaps a contrived example, but
we chose it because it requires examining the entire genotype matrix
and there is an efficient implementation available in bcftools.
It is also a good example because the standard approach of
divvying up the genome into chunks via HPC schedulers or workflow
managers is quite tricky.
We discuss the software development aspect, and how we developed
the afdist-savvy program. We discuss the inherent limitations

of the decode-to-vcf strategy, and how this inevitably
leads to bottlenecks.

\begin{figure}
\includegraphics[width=6cm]{figures/whole-matrix-compute}
\caption{Whole-matrix compute performance with increasing sample size.
(A) Total CPU time required to run \texttt{bcftools +afdist}
and equivalent operations in a single thread for various tools.
Elapsed time is also reported (dotted line). Run-time for genozip
at $10^6$ samples was extrapolated by fitting an exponential.
\label{fig-whole-matrix-compute}}
\end{figure}

>>>>

\subsection{Calculating with the genotype matrix}
Storing genetic variation data compactly is important, but it is also
important that we can analyse the data efficiently. Bioinformatics 
workflows tend to emphasise text files and command line utilities 
that consume and produce text~\citep[e.g.][]{buffalo2015bioinformatics}. 
Thus, many tools that compress VCF data provide a command line 
utility with a query language to restrict the records
examined, perform some pre-specified calculations and finally 
output some text, typically VCF or tab/comma separated values
[TODO check these citations]~\citep{
layer2016efficient, %GQT
li2016bgt, % BGT
tatwawadi2016gtrac, % GTRAC
danek2018gtc, % GTC
lin2020sparse, % SpVCF
lan2020genozip,lan2021genozip, %genozip
lefaive2021sparse, % SAVVY
wertenbroek2022xsi,% XSI
zhang2023gbc}. %GBC
These pre-defined calculations are by necessity limited in scope, however,
and the volumes of text involved in Biobank scale datasets
make the classical approach of custom
analyses via Unix utilities in pipelines prohibitively slow. Thus, 
methods have begun to provide Application Programming Interfaces
(APIs), providing efficient access to genotype and other VCF data
[TODO check these citations]~\cite[e.g.][]{lefaive2021sparse,
wertenbroek2022xsi,zhang2023gbc}.

To demonstrate the accessibility of genotype data and efficiency with 
which calculations can be performed under the different formats,
we use the \texttt{bcftools +af-dist} plugin as an example of such a whole-matrix
operation. 
% The details of the \texttt{af-dist} operation are not important:
% as an example of a whole-matrix operation. 
We chose this particular operation for several
reasons. First, it is a simple but non-trival calculation that requires 
examining every element in the genotype matrix.
Secondly, it produces a small volume of output (a table of
deviations from Hardy-Weinberg expectations in ten allele frequency
bins) and therefore the time spent outputting results is negligible.
Finally, it has an efficient implementation written using the 
\texttt{htslib} C API~\citep{bonfield2021htslib}, distributed as 
\texttt{bcftools} plugin.

\begin{figure}
\includegraphics[width=6cm]{figures/whole-matrix-compute}
\caption{Whole-matrix compute performance with increasing sample size.
Total CPU time required to run \texttt{bcftools +af-dist}
and equivalent operations in a single thread for various tools.
Elapsed time is also reported (dotted line). Run-time for genozip
at $10^6$ samples was extrapolated by fitting an exponential.
\label{fig-whole-matrix-compute}}
\end{figure}

Figure~\ref{fig-whole-matrix-compute} shows the results of 
running \texttt{bcftools +af-dist} and equivalent operations 
on the data of Figure~\ref{fig-data-storage}.

% We illustrate the discussion by the afdist operation from
% the bcftools plugins package. This is perhaps a contrived example, but
% we chose it because it requires examining the entire genotype matrix
% and there is an efficient implementation available in bcftools.
% It is also a good example because the standard approach of
% divvying up the genome into chunks via HPC schedulers or workflow
% managers is quite tricky.
% We discuss the software development aspect, and how we developed
% the afdist-savvy program. We discuss the inherent limitations
